---
title: "Biostat 203B Homework 4"
subtitle: Due ~~Mar 20~~ Mar 22 @ 11:59PM
output:
  html_document:
    toc: true
    toc_depth: 4
  # ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### 30-Day Mortality Rate of Myocardia Infarction Patients

## Introduction
MIMIC-III (Medical Information Mart for Intensive Care III) is a large, freely-available database comprising deidentified health-related data associated with over forty thousand patients who stayed in critical care units of the Beth Israel Deaconess Medical Center between 2001 and 2012.

The database includes information such as demographics, vital sign measurements made at the bedside (~1 data point per hour), laboratory test results, procedures, medications, caregiver notes, imaging reports, and mortality (both in and out of hospital).

## Data Preparation

Load database libraries and the tidyverse frontend:
```{r}
library(DBI)
library(RPostgreSQL)
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
```

Credentials for using PostgreSQL database. We are going to use username `postgres` with password `postgres` to access the `mimic` database in the schemee `mimiciii`. 
```{r}
# Load configuration settings
dbdriver <- 'PostgreSQL'
user  <- 'postgres'
password <- 'postgres'
dbname <- 'mimic'
schema <- 'mimiciii'
# Connect to the database using the configuration settings
con <- dbConnect(RPostgreSQL::PostgreSQL(), 
                 dbname = dbname, 
                 #host = host, 
                 #port = port, 
                 user = user, 
                 password = password)
# Set the default schema
dbExecute(con, paste("SET search_path TO ", schema, sep=" "))
```

```{r}
tbl(con, "d_icd_diagnoses") %>%
  filter(str_detect(tolower(long_title), "tuberculo")) %>%
  print() -> tuber_codes
```

```{r}
tbl(con, "diagnoses_icd") %>%
  semi_join(tuber_codes, by = "icd9_code") %>%
  filter(seq_num <= 5) %>%
  group_by(subject_id, hadm_id) %>%
  filter(min_rank(seq_num) <= 1) %>%
  ungroup() %>%
  select(subject_id, hadm_id, icd9_code, seq_num) %>%
  print() -> study_admissions
```


```{r}
study_admissions %>%
  mutate(principal_dx = seq_num == 1) %>%
  select(-seq_num) %>%
  print() -> study_admissions
```

```{r}
tbl(con, "drgcodes") %>%
  filter(str_detect(drg_type, "APR")) %>%
  select(subject_id, hadm_id, drg_severity) %>%
  right_join(study_admissions, by = c("subject_id", "hadm_id")) %>%
  mutate(drg_severity = ifelse(is.na(drg_severity), 1, drg_severity)) %>%
  print() -> study_admissions
```

```{r}
study_admissions %>%
  left_join(
    select(tbl(con, "admissions"),
           subject_id, hadm_id, admittime, dischtime, hospital_expire_flag
    ), by = c("subject_id", "hadm_id")
  ) %>%
  filter(hospital_expire_flag == 0) %>% # patients who did not die in hospital
  select(-hospital_expire_flag) %>%
  left_join(
    select(tbl(con, "patients"), subject_id, dob, dod),
    by = "subject_id"
  ) %>%
  print(width = Inf) -> study_admissions
```

```{r}
study_admissions %>%
  mutate(tt_death = date_part("day", dod) - date_part("day", dischtime)) %>%
  mutate(mortality = ifelse(is.na(tt_death <= 30), FALSE, tt_death <= 30)) %>%
  mutate(age = date_part("year", admittime) - date_part("year", dob)) %>%
  filter(age < 90) %>%
  mutate(age = age - ifelse(
    date_part("month", admittime) < date_part("month", dob) |
      (
        date_part("month", admittime) == date_part("month", dob) &
          date_part("day", admittime) < date_part("day", dob)
      ),
    1,
    0
  )) %>%
  select(-admittime, -dischtime, -dob, -dod, -tt_death) %>%
  select(subject_id, hadm_id, age, mortality, everything()) %>%
  print() -> study_admissions
```

```{r}
tbl(con, "admissions") %>%
  select(subject_id, ethnicity) %>%
  distinct() %>%
  print() -> study_subjects
tbl(con, "patients") %>%
  select(subject_id, gender) %>%
  distinct() %>%
  full_join(study_subjects, by = "subject_id") %>%
  print() -> study_subjects
study_subjects %>%
  semi_join(study_admissions, by = "subject_id") %>%
  print() -> study_subjects
```

```{r}
unknown_ethnicity <- c(
  "OTHER",
  "UNABLE TO OBTAIN",
  "UNKNOWN/NOT SPECIFIED",
  "MULTI RACE ETHNICITY",
  "PATIENT DECLINED TO ANSWER",
  "UNKNOWN"
)

study_subjects %>%
  collect() %>%
  mutate(ethnic_group = case_when(
    str_detect(ethnicity, "^ASIAN") ~ "ASIAN",
    str_detect(ethnicity, "^BLACK") ~ "BLACK",
    str_detect(ethnicity, "^HISPANIC") ~ "HISPANIC",
    str_detect(ethnicity, "^WHITE") ~ "WHITE",
    ethnicity %in% unknown_ethnicity ~ NA_character_,
    TRUE ~ NA_character_
  )) %>%
  select(subject_id, gender, ethnic_group) %>%
  print() -> study_subjects
```

```{r}
most <- function(x) {
  if (all(is.na(x))) return(NA_character_)
  y <- table(x, useNA = "no")
  if (length(which(y == max(y))) > 1) return(NA_character_)
  return(names(y)[which.max(y)])
}

study_subjects %>%
  group_by(subject_id) %>%
  summarize(ethnic_group = most(ethnic_group)) %>%
  ungroup() %>%
  mutate(ethnic_group = ifelse(is.na(ethnic_group), "UNKNOWN", ethnic_group)) %>%
  print() -> subject_ethnic_groups

study_subjects %>%
  select(subject_id, gender) %>%
  left_join(subject_ethnic_groups, by = "subject_id") %>%
  print() -> study_subjects
```

```{r}
study_admissions %>%
  left_join(study_subjects, by = "subject_id", copy = TRUE) %>%
  print() -> study_admissions
```
```{r}
study_admissions %>%
  distinct(.keep_all = T) %>%
  print() -> study_admissions
```

## Data Visualization

```{r}
study_admissions %>% 
  collect() %>%
  distinct(subject_id, .keep_all = T) %>%
  ggplot(aes(x = age)) + 
  geom_histogram(bins = 100) +
  labs(x = "Age", title = "Distribution of Age among Patient")
```

```{r}
study_admissions %>% 
  collect() %>%
  distinct(subject_id, .keep_all = TRUE) %>%
  ggplot(mapping = aes(x = gender, 
                       fill = gender)) + 
  geom_bar() + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.x=element_blank()) +
  geom_text(stat = 'count', aes(label = ..count.., vjust = -0.2)) +
  labs(title = "Distribution of Gender among Patient")
```
```{r}
study_admissions %>% 
  collect() %>%
  distinct(subject_id, .keep_all = TRUE) %>%
  ggplot() +
  geom_bar(mapping = aes(x = ethnic_group, fill = ethnic_group)) +
  labs(x = "Ethnic Group", y = "Counts",
       title = "Distribution of Ethnicity among Patient")
```
```{r}
study_admissions %>% 
  collect() %>%
  distinct(subject_id, .keep_all = TRUE) %>%
  ggplot(mapping = aes(x = mortality, 
                       fill = mortality)) + 
  geom_bar() + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.x=element_blank()) +
  geom_text(stat = 'count', aes(label = ..count.., vjust = -0.2)) +
  labs(title = "Distribution of Mortality among Patient")
```
```{r}
study_admissions %>% 
  collect() %>%
  ggplot(mapping = aes(x = principal_dx, 
                       fill = principal_dx)) + 
  geom_bar() + 
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.x=element_blank()) +
  geom_text(stat = 'count', aes(label = ..count.., vjust = -0.2)) +
  labs(title = "Distribution of Principal Diagnose among Admissions")
```
```{r}
study_admissions %>% 
  collect() %>%
  ggplot(mapping = aes(x = drg_severity, 
                       fill = drg_severity)) + 
  geom_bar() + 
  labs(x = "Severity Level", y = "Counts",
    title = "Distribution of Severity Level among Admissions")
```

```{r}
study_admissions %>% 
  collect() %>%
  group_by(icd9_code) %>%
  summarise(counts = n()) %>%
  arrange(desc(counts)) %>%
  slice(2:11) %>%
  ggplot(., aes(x = icd9_code, y = counts)) + 
  geom_bar(stat = 'identity') + 
  labs(x = "Diagnostics Type",
       title = "Distribution of Diagnostics Type (Top 10)")
```

## Analytics

# Split Data into a training and a validation set
```{r}
library(ISLR)
set.seed(123)
smp_siz = floor(0.75*nrow(collect(study_admissions)))
train_ind = sample(seq_len(nrow(collect(study_admissions))), size = smp_siz)  
training = collect(study_admissions)[train_ind, ] 
testing = collect(study_admissions)[-train_ind, ]
training %>%
  select(-mortality) %>%
  print() -> xtrain
training %>%
  select(mortality) %>%
  print() -> ytrain
testing %>%
  select(-mortality) %>%
  print() -> xtest
testing %>%
  select(mortality) %>%
  print() -> ytest
```
# Neural Network

Define a **sequential model** (a linear stack of layers) with 2 fully-connected hidden layers (256 and 128 neurons):
```{r}
library(keras)
model <- keras_model_sequential() 
model %>% 
  layer_dense(units = 256, activation = 'relu', input_shape = c(8)) %>% 
  layer_dropout(rate = 0.4) %>% 
  layer_dense(units = 128, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 2, activation = 'softmax')
summary(model)
```
Compile the model with appropriate loss function, optimizer, and metrics:
```{r}
model %>% compile(
  loss = 'categorical_crossentropy',
  optimizer = optimizer_rmsprop(),
  metrics = c('accuracy')
)
```
Fit the model by using the training data:
```{r}
history <- model %>% fit(
  xtrain, ytrain, 
  epochs = 15, batch_size = 128, 
  validation_split = 0.2
)
```
Evaluate model performance on the test data:
```{r}
model %>% evaluate(xtest, ytest)
```

# Logistics Regression

We fit the same model using keras, since multinomial-logit is just an MLP with (1) one input layer with linear activation and (2) one output layer with softmax link function.
```{r}
library(keras)
mlogit <- keras_model_sequential() 
mlogit %>% 
#  layer_dense(units = 256, activation = 'linear', input_shape = c(784)) %>% 
#  layer_dropout(rate = 0.4) %>% 
  layer_dense(units = 1, activation = 'softmax', input_shape = c(8))
summary(mlogit)
```
Compile the model with appropriate loss function, optimizer, and metrics:
```{r}
mlogit %>% compile(
  loss = 'categorical_crossentropy',
  optimizer = optimizer_rmsprop(),
  metrics = c('accuracy')
)
```
Fit the model by using the training data:
```{r}
mlogit_history <- mlogit %>% fit(
  xtrain, ytrain, 
  epochs = 20, batch_size = 128, 
  validation_split = 0.2
)
```
Evaluate model performance on the test data:
```{r}
mlogit %>% evaluate(xtest, ytest)
```


## Conclusion
